\section{Evaluation}

We organize the evaluation around four program-level research questions---RQ1 (audited Bugs4Q defects), RQ2 (balanced injected positive control), RQ3 (false positives), and RQ4 (sensitivity floor)---plus a circuit-level-only diagnostic run (\S\ref{sec:diagnostic}). Each claim is tied to a dataset, baseline, and metric, and each dataset is introduced with its research question below. MorphQ is the only matched baseline; QMutPy is not evaluated (it assumes an existing test suite), and stack/assertion tools are related-only comparators (Table~\ref{tab:related-comparison}). The primary evidence is program-level; the circuit-level run is retained only as a diagnostic and is not used as a main claim.

\subsection{Execution and Evidence Pipeline}

All runs use Qiskit Aer under a pinned environment, a fixed master seed, a default shot budget for detection, and a per-pair wall-clock timeout. MorphQ is run on the same subjects under identical backends, shot budgets, seeds, and timeouts. Every headline number resolves to one row of a canonical CSV that is derived by \texttt{derive\_metrics.py} from run-isolated summary shards under a frozen run directory; the derivation enforces expected denominators, rejects missing or stale-mixed rows, and re-derives the canonical CSV to a byte-identical copy (\emph{canonical integrity check: PASS}). Table~\ref{tab:canonical-results} is the canonical view.

\input{tables/canonical_results}

\subsection{RQ1: Real Program-Level Defects}
\label{sec:rq1}

\noindent\textbf{How many audited Bugs4Q program-level defects does QMTester detect?}

We \emph{included} a Bugs4Q subject only when a sound program-level relation could be justified (an input transform, a bijective canonicalization map, and a soundness argument tied to builder semantics). The denominators form one \emph{closed} chain: of the 42 Bugs4Q bugs~\cite{miao2023bugs4q}, 32 build and run under our pinned Qiskit (the diagnostic set, \S\ref{sec:diagnostic}); we audited \emph{all 32} (per-bug dispositions in the artifact), of which \emph{9} admit a sound builder-contract relation and 23 are out of scope (12 api-misuse, the rest output-formatting, platform/runtime, or wrong-gate; Table~\ref{tab:excluded-bugs4q}). Seven of the nine are runnable distribution-family subjects (Table~\ref{tab:audited-bugs4q}), all detected; the \texttt{program\_register\_shape} static check flags the eighth---\texttt{qiskit\_github\_12}, the \texttt{measure\_all} defect previously excluded for lacking a bijective map---and the ninth (a QPE/inverse-QFT case) awaits a sound relation. QMTester thus \emph{handles 8} of the 9 admitted defects (7/7 on the distribution families) and flags none of the matched fixed variants. Beyond Bugs4Q, mining recent Quantum Computing StackExchange (2022--2024) adds \emph{three} verified real defects (two QFT-round-trip, one input-permutation), each held to the \emph{same} bar---buggy detected, matched fix clean (\texttt{program\_mined\_manifest}). The real distribution base is thus \emph{ten}, all detected at $0$ false positives (Wilson $[0.722,1.000]$, up from $[0.646,1.000]$). The matched baselines extend: the reference-oracle detects all three (so QMTester matches the ceiling with \emph{no} oracle on all \emph{ten}), the static analyzer none ($0/10$); a circuit-level diagnostic incidentally fires on the QPE via swap-sensitivity (not a builder-contract detection, \S\ref{sec:diagnostic}).

\input{tables/audited_bugs4q}
\input{tables/excluded_bugs4q}

\noindent This is detection on the relation-applicable subset, not a suite-wide rate. Three matched baselines (Table~\ref{tab:program-baselines}): MorphQ is structurally blind ($0$); the \emph{deployable} static analyzer (LintQ-style register-shape check) also detects $0$---the circuits being equally well-formed, it flags only the one \emph{syntactic} \texttt{measure\_all} case (covered by \texttt{program\_register\_shape}); a reference-oracle that \emph{has} each known-correct version detects all, the ceiling but not deployable. QMTester reaches that ceiling with \emph{no} oracle; MorphQ's $0$ is circuit-level blindness---structural on the builder-only families, a catalog gap on the two circuit-expressible ones (\S\ref{sec:relations}). We are plain about what the statistics do \emph{not} carry: detection on all ten (deterministic) real subjects is carried by canonicalization (canonicalize-plus-raw-equality reaches $10/10$, \S\ref{sec:oracle-light}). The permutation test and Bonferroni earn their keep on \emph{specificity} and \texttt{parameter\_periodicity}, the one probabilistic family, which still has \emph{no real subject} (only the synthetic power curve, Table~\ref{tab:power-curve}).

\subsection{RQ2: Injected Positive Control}

\noindent\emph{Sanity check (coverage-by-construction).} Does each family fire on a balanced injected control, and what does QMTester's machinery add over a trivial rebuild-and-compare?

Because a mutant is admitted here only if detected (\S\ref{sec:threats}), the $100/100$ is coverage-by-construction (Table~\ref{tab:canonical-results}): a sanity check that each family fires on its own contract, \emph{not} a detection rate or contest. We retain it only as the within-method ablation substrate below. (MorphQ detects $20/100$---injected single-circuit \texttt{ancilla\_uncompute} mutants whose garbage \emph{is} circuit-visible---but \emph{none} of the three \emph{real} cross-boundary ancilla subjects, so its blindness holds for the real defects, not the injected proxy.)
\label{sec:oracle-light}

\noindent\emph{Oracle-light baselines.} To isolate QMTester's machinery, three \emph{program-level} baselines share the rebuild but each strip one component (Table~\ref{tab:program-baselines}): \emph{same input} (no transform), \emph{no canonicalization} (raw supports to the oracle), and \emph{raw equality}---same seven subjects and 100 mutants, identical shots/seed/correction.

\input{tables/program_baselines}

The result isolates two contributions. First, the metamorphic transform is necessary: \emph{same input} detects $0/7$ and $0/100$. Second, canonicalization buys specificity: dropping it makes \emph{no canonicalization} and \emph{raw equality} false-positive on correct variants ($2/7$, $40/100$) \emph{and} miss buggy ones ($5/7$, $60/100$ detected), the un-undone representation change both masquerading as a fault and masking real divergences. The real-subject differential is exactly the two \texttt{classical\_remap} cases (2 points); the broader, inclusion-independent signal is the $40/100$ injected fixed-variant false positives QMTester's full pipeline holds at $0$---within-method evidence MorphQ cannot provide.

\subsection{RQ3: False Positives}

\noindent\textbf{What false-positive rate does QMTester report?}

Specificity is the only inclusion-\emph{independent} evidence in our evaluation, so we report it carefully. Across the 107 fixed program variants (7 Bugs4Q fixed subjects, 100 injected fixed variants) and the 50-program correct corpus, QMTester raises $0$ false positives under the ideal, noisy, and calibrated-noisy simulators (Table~\ref{tab:canonical-results}). The \emph{ideal} cell is degenerate (matched seeds make source and follow-up bit-identical, $p\equiv 1$); the informative cells are the noisy ones. Under the bundled depolarizing model ($p_1{=}10^{-3}$, $p_2{=}5{\times}10^{-3}$, $T_1{=}T_2{=}80\,\mu$s, $t_g{=}50$\,ns, density-matrix sim; \texttt{noise.json}) the smallest \emph{uncorrected} per-pair $p$ over all $3233$ pairs is $0.370$ (noisy) and $0.251$ (calibrated)---conservative, but not by a wide margin. To stress this we re-ran the corpus under a \emph{realistic} model adding $2\%$ symmetric readout error and a $1\%$ two-qubit crosstalk term: the smallest uncorrected $p$ falls to $0.0035$ (below the nominal $0.05$), yet source-level Bonferroni still yields $0/50$. So under realistic noise the per-pair margin is genuinely thin and the multiple-testing correction---not merely conservative nulls---is what holds specificity, a qualifier the four $0/50$ rows alone would hide. Concretely, the thin-margin subject (smallest realistic-noise $p=0.0035$) carries $m=96$ admitted pairs, so $\alpha/m=0.0005\le0.0035$; across the corpus $m\in[30,96]$, every per-subject threshold below its subject's smallest margin, so $0/50$ holds by correction, not by the thin-margin pair landing on a high-$m$ subject by luck.

\subsection{Statistical-Oracle Behavior}
\label{sec:oracle-behavior}

We verify two properties of the statistical oracle directly from the logged per-pair records. \emph{Empirical type-I:} across the 50 correct programs, none of the $3233$ per-pair null p-values falls below $0.05$ even uncorrected under any of the three simulators (minimum $1.000$/$0.370$/$0.251$ for ideal/noisy/calibrated-noisy). The nulls are thus not uniform but \emph{conservative}, the favorable direction for Remark~1, matching the observed $0/107$ and $0/50$ false-positive rates. \emph{Which branch decides:} the Monte-Carlo permutation branch, not the asymptotic $\chi^2$, fires for most program-level decisions ($85.7\%$ of Bugs4Q and $93.0\%$ of injected pairs), as the small quantum supports never satisfy the Cochran asymptotic condition (zero merges across all admitted pairs); we therefore describe the oracle as permutation-primary with an asymptotic $\chi^2$ fast path.

\subsection{RQ4: Sensitivity Floor (Detection vs.\ Effect Size)}

\noindent\textbf{At what effect size does detection start to fail?}

The families differ in output character. \texttt{input\_permutation}, \texttt{classical\_remap}, \texttt{ancilla\_uncompute}, and the real \texttt{qft\_round\_trip} subject all produce \emph{deterministic} basis-state divergence (TVD$=1.0$); the $0.74$--$0.94$ figure is the input-dependent effect size of the \emph{injected} QFT mutants, not the real subject. Only the probabilistic \texttt{parameter\_periodicity} family has a tunable effect. A $204$-mutant rotation-drift sweep (TVD$\in(0,0.06)$, \emph{not} filtered by detectability; true effect at 32{,}768 shots, detection at 4096) localizes the threshold via a fine sub-$0.03$ grid (Table~\ref{tab:power-curve}): the $50\%$/$80\%$ crossings fall in $[0.022,0.025)$/$[0.025,0.030)$. So the conservative nulls do not silently trade away sensitivity---the method detects reliably above a measured floor near TVD~$0.025$ (probabilistic family only).

\input{tables/power_curve}

\subsection{Diagnostic Circuit-Level Run}
\label{sec:diagnostic}

For transparency, a circuit-level-only configuration (identity/swap/phase/equivalence rewritings) shows \emph{no} advantage over MorphQ---$2/32$ on Bugs4Q, $17/250$ injected, identical for both, all from swap---motivating the program-level design (not a contribution claim).
